---
title: "Assignment 2 - Social Network Analysis Solution"
output: html_document
author: Shreya Goel
---
# Libraries
```{r}
library(tidyr)
library(dplyr)
library(igraph)
```


# Data Wrangling
```{r}
#Read data into
D1 <- read.csv("hudk4050-classes.csv", header = TRUE)

#Merge First.name and Last.name variables to create unique ID because we have duplicate first and last names in the class
D1 <- tidyr::unite(D1, Name, Last.name, First.name, sep = " ", remove = TRUE)

#Reshape data to create a "course" variable (you will get a warning because there are missing cells)
D2 <- tidyr::gather(D1, course.label, course, Course1, Course2, Course3, Course4, Course5, na.rm = TRUE, convert = FALSE)
#or 
#D_long <- tidyr::gather(D1, CourseNo., "course", 2:6)

#Remove the "course.label" variable
D2 <- dplyr::select(D2, Name, course)
#or
#D2 <- D_long[c(1,3)]

#Remove rows indicating HUDK4050 because all students are in this course and it will overwhelm the graph
D2 <- dplyr::filter(D2, course > 0, course != "HUDK4050")

#Add a variable to be used to count the courses
D2$Count <- 1

#Reshape the data to create a person x class matrix
D3 <- tidyr::spread(D2, course, Count)

#This was a bit of a trick, for the matrix command to work the row names needed to changed from an indice (1,2,3,etc) to the student names 
row.names(D3) <- D3$Name
D3$Name <- NULL

D3 <- ifelse(is.na(D3), 0, 1)

#Convert the data from data frame format to matrix format so it can be transposed
D4 <- as.matrix(D3)

#Transpose matrix to produce a person x person matrix
D5 <- D4 %*% t(D4) #to get the number of courses these people are doing together
diag(D5) <- NA #to remove the diagnal - the number of courses a person is doing with self. Redundant
#Transpose. The transpose of a matrix is a new matrix whose rows are the columns of the original. (This makes the columns of the new matrix the rows of the original)
```
# Graphing
```{r}
g <- graph.adjacency(D5,mode="undirected")

plot(g,layout=layout.fruchterman.reingold, vertex.size=6)

```
# Centrality
```{r}
#Calculate the degree centrality of the nodes, showing who has the most connections
degree(g)

#Calculate the betweeness centrality, showing how many "shortest paths" pass through each node. This turns out to be uniformative for this graph as there are a bunch of people who are almost equally connected and a bunch of people who are unconnected.
betweenness(g)
```







TO clear console -> ctrl + L

```{r}
getwd()
```

```{r}
install.packages("igraph")
library(igraph)
```

Now upload the data file "discipline-data.csv" as a data frame called "D1". Each row is a disciplinary action from a teacher to a student so the first line shows that teacher "E" sent student "21" to the principal. It also shows the gender of both the teacher and student and the student's main elective field of study ("major"") and the field that the teacher instructs in ("t.expertise").


```{r}
getwd()
```

```{r}
D1 <- read.csv(".../discipline-data.csv")
```

Before you proceed, you will need to change the data type of the student id variable. Since it is a number R will automatically think it is an integer and code it as such (look at the list of variables by clicking on the data frame arrow in the Data pane. Here you will see the letters "int"" next to the stid variable, that stands for integer). However, in this case we are treating the variable as a category, there is no numeric meaning in the variable. So we need to change the format to be a category, what R calls a "factor". We can do this with the following code:

```{r}
D1$stid <- as.factor(D1$stid)
```

igraph requires data to be in a particular structure. There are several structures that it can use but we will be using a combination of an "edge list" and a "vertex list". As you might imagine the edge list contains a list of all the relationships between students and teachers and any characteristics of those edges that we might be interested in. There are two essential variables in the edge list a "from" variable and a "to" variable that descibe the relationships between vertices (a disciplinary action is given "from" and teacher "to" a student). While the vertix list contains all the characteristics of those vertices, in our case gender and major.

So let's convert our data into an edge list!

First we will isolate the variables that are of interest: tid and stid

```{r}
install.packages("dplyr", dependencies = TRUE)
library(dplyr)

D2 <- dplyr::select(D1, tid, stid)
```

Since our data represnts every time a teacher sends a student to the principal there are multiple rows when the same teacher sends the same student. We want to collapse these into a single row, with a variable that shows how many times a teacher-student pair appears.

```{r}
EDGE <- dplyr::count(D2, tid, stid)
```

```{r}
names(EDGE) <- c("from", "to", "count")
```

EDGE is your edge list. Now we need to make the vertex list, a list of all the teachers and students and their characteristics in our network.

```{r}
#First we will separate the teachers from our original data frame
V.TCH <- dplyr::select(D1, tid, t.gender, t.expertise)
```

```{r}
#Remove all the repeats so that we just have a list of each teacher and their characteristics
V.TCH <- unique(V.TCH)
```

```{r}
#Add a variable that describes that they are teachers
V.TCH$group <- "teacher"
```

```{r}
#Now repeat this process for the students
V.STD <- dplyr::select(D1, stid, s.gender, s.major)
V.STD <- unique(V.STD)
V.STD$group <- "student"
```

```{r}
#Make sure that the student and teacher data frames have the same variables names
names(V.TCH) <- c("id", "gender", "topic", "group")
names(V.STD) <- c("id", "gender", "topic", "group")
```

```{r}
#Bind the two data frames together (you will get a warning because the teacher data frame has 5 types of id (A,B,C,D,E) and the student has 25 (1-30), this isn't a problem)
VERTEX <- dplyr::bind_rows(V.TCH, V.STD)
```

Now we have both a Vertex and Edge list it is time to plot our graph!

```{r}
#Load the igraph package

library(igraph)

#First we will make an object that contains the graph information using our two dataframes EDGE and VERTEX. Notice that we have made "directed = TRUE" - our graph is directed since discipline is being given from a teacher to a student.

g <- graph.data.frame(EDGE, directed=TRUE, vertices=VERTEX)
```

```{r}
#Now we can plot our graph using the force directed graphing technique - our old friend Fruchertman-Reingold!

plot(g,layout=layout.fruchterman.reingold) #Rplot1
```

```{r}
#There are many ways to change the attributes of the graph to represent different characteristics of the newtork. For example, we can color the nodes according to gender.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender)
```

```{r}
#We can change the thickness of the edge according to the number of times a particular teacher has sent a particular student to the principal.

plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender, edge.width=EDGE$count)
```

```{r}
#Other attributes you may want to change are:

#Arrow size
# -> edge.arrow.size=

#Remove the labels
# -> vertex.label=NA

#Vertex size
# -> vertex.size=10
  
#And many others!
  
plot(g,layout=layout.fruchterman.reingold, vertex.color=VERTEX$gender, edge.width=EDGE$count, edge.arrow.size=0.3)
```

Play around with different graphs and then save your final version to the repo. Create a Zotero item for this R program and then commit, push and pull.

Once you have done this you need to practice with data from our class. Please create a **person-network** with the data set hudk4050-classes.csv. Once you have done this, also [look up](http://igraph.org/r/) how to generate the following network metrics: betweeness centrality and dregree. **Who is the most central person in the network?**

```{r}
install.packages("igraph")
install.packages("tidyr")
install.packages("dplyr")
```

```{r}
getwd()
```

```{r}

library(igraph)
library(tidyr)
library(dplyr)


P1 <- read.csv("C:/Users/Shreya/Documents/GITHUB/4050_2_SNA/hudk4050-classes.csv", header = TRUE)
FullName <- as.data.frame(P1)

P1$FullName <- do.call(paste, c(P1[c("First.name", "Last.name")], sep = " "))

library(dplyr)
C1 <- dplyr::select(P1, FullName, Course1, Course2, Course3, Course4, Course5)

library(tidyr)
C_long <- tidyr::gather(C1, CourseNo., "Courses", 2:6)

names(C_long) <- c("Names", "CourseNo.", "Courses")

library(dplyr)
C2 <- dplyr::select(C_long, Names, Courses)
C2$Courses <- ifelse(C2$Courses == "", NA, C2$Courses)

C3 <- na.omit(C2)

Table1 <- table(C3$Names, C3$Courses)

C5 <- as.data.frame.table(Table1)
names(C5) <- c("from", "to", "count")
C5$count <- ifelse(C5$count == "0", NA, C5$count)

C6 <- na.omit(C5)

V1 <- dplyr::select(C3, Courses)
V1 <- unique(V1)
V1 <- dplyr::arrange(V1, Courses)
#this aggranges fields in alphabetical order
V1$group <- "course"

V2 <- dplyr::select(C3, Names)
V2 <- unique(V2)
V2$group <- "Student"

names(V1) <- c("id",  "group")
names(V2) <- c("id", "group")

V <- dplyr::bind_rows(V1, V2)

V$group <- ifelse(V$group == "course", 100, 1)

g <- graph.data.frame(C6, directed=TRUE, vertices=V)
plot(g,layout=layout.fruchterman.reingold, edge.arrow.size=0.3, vertex.color=V$group)
```

